Measurement Bias from Address Aliasing

### 2014-05-25 paper.tex
Formatting for tables, figures and listings

% Typeset for Transactions on Architecture and Code Optimization
\documentclass[prodmode,acmtaco]{acmsmall}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{filecontents}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{listings}
\usepackage{caption}
\usepackage{mathtools}
\usepackage{underscore}
\usepackage{url}
\usepackage{lmodern} % Much clearer text

@@

% Generic intro to memory layout (with focus on stack/env interaction)
% todo: Consider removing?
Figure \ref{fig:virtualmemory} shows the relative positions of some important sections of memory at runtime, assuming a 64-bit process mapped to virtual memory. 
The stack is placed close to the upper address 0x7fff'ffffffff\footnote{Modern processors do not actually use the full 64-bit space, only the low order 47 bits are used for addressing memory}, but offset by environment variables and program arguments.
After some variable amount of offset, the stack is normally aligned to a 16 byte boundary, which will be enforced by the compiler. % [ref]
Within a span of 4096 bytes there are thus 256 possible initial addresses, each representing a different execution context.
Any one of these contexts could trigger aliasing issues between stack and some other area of memory remaining unaffected by environment size.
This includes dynamic memory allocateded on heap, memory mapped files, as well as statically allocated data. 
% Note that there is no clear relationship between environment size and stack location with ASLR enabled. % todo: check again
% However, there will still be as many execution contexts with respect to aliasing (considering stack only), making any occurrences of measurement bias indeed random.

\begin{figure}[t]
  \centering
  \caption{Organization of virtual memory for a 64-bit process.}
  \label{fig:virtualmemory}
  \begin{tikzpicture}[font=\footnotesize]

@@

  \end{tikzpicture}
\end{figure}

\subsection{Microkernel Analysis}
To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!''~\cite{Mytkowicz:2009:WrongData}, reproduced here in Listing~\ref{lst:loopkernel}. 
This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the example code is simple and straightforward to analyze.
Still, no satisfactory explanation as to what actually causes bias was given in the original paper.

% Cite does not work in acm style
% Micro-kernel presented by Mytkowicz et al~\cite{Mytkowicz:2009:WrongData}, 
\begin{lstlisting}[float=h, language=C, caption={Micro-kernel succeptible to bias from variations to system environment variables.}, label={lst:loopkernel}, frame=lines]
static int i, j, k;
int main() {
    int g = 0, inc = 1;
    for (; g < 65536; g++) {
        i += inc;
        j += inc;
        k += inc; 
    }
    return 0;
}
\end{lstlisting}

% Large plot of cycle count for two 4K stack sections.
\pgfplotstableread{bin/micro-kernel-cycles.dat}{\stackoffsettable}
\begin{figure*}[t]
  \caption{Bias from environment size for micro-kernel in Listing~\ref{lst:loopkernel}. Measured average of 10 cycle count samples for 512 different environments. Spikes show aliasing case, occurring once for each 4K period}
  \label{fig:envbias}
  \begin{tikzpicture}
    \begin{axis}[
        width = \textwidth,
        height = 6cm,
        font = \footnotesize,
        xlabel=Bytes added to environment,
        ylabel=Cycles,
        domain = 0:8192,
        xtick = {0,1024,...,8192},
        xmin = 0,
        xmax = 8192,
        cycle list name = exotic
      ]
      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
    \end{axis}
  \end{tikzpicture}
\end{figure*}

% Detailed comparison between median and worst cases.
\begin{table}[b]
  \centering
  \caption{Events with significant correlation to cycle count}
  \label{tab:loopcorrelation}
  %\renewcommand{\tabcolsep}{1pt} % Less padding between cells
  \renewcommand{\tabcolsep}{4pt} % Less padding between cells
  \pgfplotstabletypeset[
    font=\footnotesize,
    int detect, % Output whole numbers for counter values
    col sep=comma,
    columns={Performance counter, Median, [index]3, [index]4},
    column type=r,
    columns/Performance counter/.style={
      string type, 
      column type=l,
      postproc cell content/.append code={
        \pgfkeysalso{@cell content=\perfctr{##1}}
      }
    },
    every head row/.style={
      output empty row,
      before row={\toprule
        Performance counter & Median & Spike 1 & Spike 2 \\
        },
      after row=\midrule,
    },
    every last row/.style={after row=\bottomrule}
  ]{bin/micro-kernel-comparison.csv}
\end{table}

@@

@@

% Refer to relevant section of the generated assembly file (in gas syntax)
\lstinputlisting[
  linerange={22-38}, 
  numbers=left,
  firstnumber=22
  ]{bin/micro-kernel-annotated.s}

@@

\subsection{Avoiding Aliasing}
Addresses of automatic variables can not be determined statically, because the position of stack at runtime is generally unknown. 
In addition to being offset by environment variables, the stack address can also be perturbed by other factors such as ASLR. 
Although one can not easily know if a collision is going to happen for a given environment, it is possible to change the program to account for possible alias effects.
Shown in Listing \ref{lst:loopfixed} is a proof of concept of how alias-free code could be generated in this particular case.
If the addresses do alias, a branch to an alternative but semantically equivalent code path is performed.
Calling the function recursively will effectively allocate a new set of variables a bit further down the stack, and the alias condition is avoided.

% See analysis for results showing this does not alias
\begin{lstlisting}[float=t, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopfixed}, frame=lines]
#define ALIAS(a, b) \
    (((long)&a) & 0xfff == ((long)&b) & 0xfff)
static int i, j, k;
int main() {
    int g = 0, inc = 1;
    if (ALIAS(inc, i) || ALIAS(g, i))
        return main();
    for (; g < 65536; g++) {
        i += inc;
        j += inc;
        k += inc;
    }
    return 0;
}
\end{lstlisting}

@@

\subsection{\texttt{malloc} will Alias by Default}
Acquiring dynamic memory at runtime is usually done by calling \texttt{malloc}, which takes a number of bytes to allocate as input, and returns a pointer to that area.
A typical malloc implementation uses different strategies for how to allocate memory depending on the \emph{size} of the request.
\begin{itemize}
  \item The ``regular'' heap area in virtual memory is used for smaller allocations. 
  Programs are initially given a relatively small heap, which the allocator can increase by calling \texttt{brk} or \texttt{sbrk} system calls.

  \item For larger allocations, it can be more efficient to use anonymous memory mapping by calling the Posix function \texttt{mmap} instead of growing the regular heap.
  One drawback of this is that it requires the memory to be zero initialized, and whether memory mapping is used or not is based on some cost heuristic employed by the allocator.
\end{itemize}
The heap section starts at a relatively low address (close to the data section), growing upwards.
Memory mapped chunks are placed towards the upper end of the virtual address space, closer to the stack.
Whether a memory allocation request is served by the heap or by memory mapping is therefore easy to determine just by looking at the actual pointer values returned.
While addresses in the regular heap can look something like 0x16e30a0 or 0x1723020, pointers returned by \texttt{mmap} are numerically much larger -- for example 0x7f0318a8f010 or 0x7f03105d2010.
This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
However, \texttt{mmap} has an interesting property in that allocations will always be page aligned.
The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of metadata at the beginning, therefore every memory mapped address end with 0x010.}
This behavior is often the worst case for functions that operate on two or more independent buffers, as we will see in the following sections.

\subsection{Aligned Sequential Access}
Many functions operate in a ``sliding window'' fashion; reading and writing to different buffers in some loop construction.
This type of program is potentially vulnerable to 4K aliasing, where the worst case will be when the read and write pointer addresses are aliased, continuously generating false conflicts.
As an example of this, consider a naive implementation of linear correlation shown in Listing \ref{lst:conv} (for simplicity, endpoints are not handled).
The function containins lots of interleaved loads and stores, iterating over two independent memory buffers in a tight loop.
We will see that the performance of this program greatly depends on the address alignment of each buffer, favoring memory addresses that are not closely aligned on the last 12-bits.

\lstinputlisting[
  float=t, 
  language=C, 
  caption={Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.},
  label={lst:conv},
  frame=lines
]{bin/convolution-kernel.c}

% Use mmap directly instead maybe?
By using an input size of $n=2^{20}$ (4 MiB in memory for each array), glibc's heap allocator ends up always choosing \texttt{mmap} to serve requests.
By default, even with address randomization enabled, both pointers will have the same address suffix of 0x010.
To analyse performance for different addresses, we manually insert padding to offset one of the memory mapped pointers.
This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset one of the arguments before calling the function.
Controlling the offset parameter, we can create environments where the inputs are some number of \texttt{sizeof(float)} bytes apart.
Allocating and managing these buffers at startup takes a non-neglegible amount of work.
The overhead can be masked by repeatedly invoking the \texttt{convolution} kernel after allocating and initializing all the inputs.
\begin{lstlisting}
    for (i = 0; i < k; ++i)
        convolution(n, input, output + offset);
\end{lstlisting}
An estimate of the actual cost of one invocation can be calculated by averaging over a number of repeated function calls, after subtracting the constant overhead of invoking it only once.

@@

\pgfplotstableread{bin/conv-default-o2.estimate.dat}{\convtabletwo}
\pgfplotstableread{bin/conv-default-o3.estimate.dat}{\convtablethree}
\begin{figure*}[t]
  \centering
  \caption{Estimated cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Listing~\ref{lst:conv}. Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations. Input size $n=2^{20}$.}
  \label{fig:conv-default}
  \begin{tikzpicture}
    \begin{axis}[
        title=\texttt{cc -O2},
        font=\footnotesize,
        xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
        cycle list name=black white,
        width=\textwidth/1.9, % Make it fit to text width side by side
        skip coords between index={20}{32} % Limit level of detail to fit page width nicely
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtabletwo ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwo ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=\texttt{cc -O3},
        font=\footnotesize,
        xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
        cycle list name=black white,
        width=\textwidth/1.9,
        skip coords between index={20}{32}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtablethree ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtablethree ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
\end{figure*}

% Table with detailed performance data, use estimated for O2 because it does not need as many columns (!)
\begin{table*}[t]
  \centering
  \caption{Most interesting performance counters and correlation with cycle count for optimization O2, estimated cost accounting for constant overhead.}
  \label{tab:convstats}
  \renewcommand{\tabcolsep}{4pt} % Less padding between cells
  \pgfplotstabletypeset[
    font=\footnotesize,
    int detect, % Output whole numbers for counter values
    col sep=comma,
    columns={Performance counter, Correlation, 0, 2, 4, 8},
    column type=r,
    columns/Performance counter/.style={
      string type, 
      column type=l,
      postproc cell content/.append code={
        \pgfkeysalso{@cell content=\perfctr{##1}}
      }
    },
    columns/Correlation/.style={
      fixed,
      fixed zerofill,
      precision=2
    },
    every head row/.style={
      output empty row,
      before row={\toprule
        Performance counter & $r$ & 0 & 2 & 4 & 8 \\
        },
      after row=\midrule,
    },
    every last row/.style={after row=\bottomrule}
  ]{bin/conv-default-o2.estimate.csv}
\end{table*}

@@

\begin{table*}[t]
  \centering
  \caption{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.}
  \label{tab:mallocompare}
  \renewcommand{\tabcolsep}{4pt} % Less padding between cells
  \pgfplotstabletypeset[
    font=\footnotesize,
    col sep=comma,
    string type,
    columns={Allocation, 64, 5120, 65536, 1048576, 4194304},
    column type=r,
    columns/Allocation/.style={
      string type, 
      column type=l
    },
    every head row/.style={
      output empty row,
      before row={\toprule
         & 64 B & 5,120 B & 65,536 B & 1,048,576 B & 4,194,304 B \\
        },
      after row=\midrule,
    },
    every odd row/.style={after row=\midrule},
    every last row/.style={after row=\bottomrule}
  ]{bin/malloc-comparison.csv}
\end{table*}

@@

\paragraph{Manually adjust memory}
In some cases it might make sense to explicitly control the memory addresses used, for example forcing some fixed relative offset between input and output pointers.
This can be achieved by exploiting the fact the \texttt{mmap} is in fact guaranteed to be placed at a page boundary, and use that directly instead of \texttt{malloc}.
The following approach can be used to make an anonymous memory mapping with offset \texttt{d} bytes away from page alignment.
\begin{lstlisting}[breaklines=true]
    mmap(NULL, (n + d), PROT_READ | PROT_WRITE, MAP_PRIVATE | MAP_ANONYMOUS, -1, 0) + d;
\end{lstlisting}
The same pointer difference must of course by subtracted before unmapping the memory.

>>>>

% Typeset for Transactions on Architecture and Code Optimization
\documentclass[prodmode,acmtaco]{acmsmall}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{filecontents}
\usepackage{booktabs} % Only used for horizontal lines
\usepackage{tikz}
\usepackage{listings}
%\usepackage{caption}
\usepackage{mathtools}
\usepackage{underscore}
\usepackage{url}
\usepackage{lmodern} % Much clearer text

@@

\begin{narrowfig}{0.5\textwidth}[t]
  \centering
  \begin{tikzpicture}[font=\footnotesize]

@@

  \end{tikzpicture}
  \caption{Relative positions of some important sections of memory at runtime, assuming a 64-bit process mapped to virtual memory. 
  The stack is placed close to the top, but offset by environment variables and program arguments. 
  Program code and statically allocated data is allocated at the lower end of the address space.
  The memory addresses of these sections are at compile time by the linker, and can be determined by inspecting the executable.}
  % \footnote{Modern processors do not actually use the full 64-bit space, only the low order 47 bits are used for addressing memory}
  \label{fig:virtualmemory}
\end{narrowfig}

% Generic intro to memory layout (with focus on stack/env interaction)
% todo: Consider removing?

% Figure \ref{fig:virtualmemory} shows the relative positions of some important sections of memory at runtime, assuming a 64-bit process mapped to virtual memory. 
% The stack is placed close to the upper address 0x7fff'ffffffff\footnote{Modern processors do not actually use the full 64-bit space, only the low order 47 bits are used for addressing memory}, but offset by environment variables and program arguments.
% After some variable amount of offset, the stack is normally aligned to a 16 byte boundary, which will be enforced by the compiler. % [ref]
% Within a span of 4096 bytes there are thus 256 possible initial addresses, each representing a different execution context.
% Any one of these contexts could trigger aliasing issues between stack and some other area of memory remaining unaffected by environment size.
% This includes dynamic memory allocateded on heap, memory mapped files, as well as statically allocated data.

% Note that there is no clear relationship between environment size and stack location with ASLR enabled. % todo: check again
% However, there will still be as many execution contexts with respect to aliasing (considering stack only), making any occurrences of measurement bias indeed random.

\subsection{Microkernel Analysis}
To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!''~\cite{Mytkowicz:2009:WrongData}, reproduced below. 
This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the example code is simple and straightforward to analyze.
Still, no satisfactory explanation as to what actually causes bias was given in the original paper.

% Cite does not work in acm style
%\begin{figure}
  % \lstset{%
  %   basicstyle=\ttfamily\small\bfseries,
  %   frame=tb,
  %   xleftmargin=0.2\textwidth,
  %   xrightmargin=0.2\textwidth
  % }
  % [xleftmargin=0.2\textwidth,xrightmargin=0.2\textwidth]
  \begin{lstlisting}
    static int i, j, k;
    int main() {
        int g = 0, inc = 1;
        for (; g < 65536; g++) {
            i += inc;
            j += inc;
            k += inc; 
        }
        return 0;
    }
  \end{lstlisting}

  % Can have caption with figure environment
  %\caption{Micro-kernel succeptible to bias from variations to system environment variables.\label{lst:loopkernel}}
%\end{figure}

% Large plot of cycle count for two 4K stack sections.
\pgfplotstableread{bin/micro-kernel-cycles.dat}{\stackoffsettable}
\begin{figure*}[t]
  \begin{tikzpicture}
    \begin{axis}[
        width = \textwidth,
        height = 6cm,
        font = \footnotesize,
        xlabel=Bytes added to environment,
        ylabel=Cycles,
        domain = 0:8192,
        xtick = {0,1024,...,8192},
        xmin = 0,
        xmax = 8192,
        cycle list name = exotic
      ]
      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
    \end{axis}
  \end{tikzpicture}
  \caption{Bias from environment size for microkernel. Measured average of 10 cycle count samples for 512 different environments. Spikes show aliasing case, occurring once for each 4K period}
  \label{fig:envbias}
\end{figure*}

@@

% Detailed comparison between median and worst cases.
\begin{table}
  \tbl{Events with significant correlation to cycle count\label{tab:loopcorrelation}}{
    \pgfplotstabletypeset[
      int detect, % Output whole numbers for counter values
      col sep=comma,
      columns={Performance counter, Median, [index]3, [index]4},
      column type=r,
      columns/Performance counter/.style={
        string type, 
        column type=l,
        postproc cell content/.append code={
          \pgfkeysalso{@cell content=\perfctr{##1}}
        }
      },
      every head row/.style={
        output empty row,
        before row={\toprule
          Performance counter & Median & Spike 1 & Spike 2 \\
          },
        after row=\midrule,
      },
      every last row/.style={after row=\bottomrule}
    ]{bin/micro-kernel-comparison.csv}
  }
\end{table}

@@

% Refer to relevant section of the generated assembly file (in gas syntax)
\lstinputlisting[
  basicstyle=\ttfamily\footnotesize,
  linerange={22-38}, 
  numbers=left,
  firstnumber=22
  ]{bin/micro-kernel-annotated.s}

@@

\subsection{Avoiding Aliasing}
Addresses of automatic variables can not be determined statically, because the position of stack at runtime is generally unknown. 
In addition to being offset by environment variables, the stack address can also be perturbed by other factors such as ASLR. 
Although one can not easily know if a collision is going to happen for a given environment, it is possible to change the program to account for possible alias effects.
Shown in Figure \ref{lst:loopfixed} is a proof of concept of how alias-free code could be generated in this particular case.
If the addresses do alias, a branch to an alternative but semantically equivalent code path is performed.
Calling the function recursively will effectively allocate a new set of variables a bit further down the stack, and the alias condition is avoided.

% See analysis for results showing this does not alias
\begin{figure}
  % [float=t, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopfixed}, frame=lines]
  \begin{lstlisting}
  #define ALIAS(a, b) \
      (((long)&a) & 0xfff == ((long)&b) & 0xfff)
  static int i, j, k;
  int main() {
      int g = 0, inc = 1;
      if (ALIAS(inc, i) || ALIAS(g, i))
          return main();
      for (; g < 65536; g++) {
          i += inc;
          j += inc;
          k += inc;
      }
      return 0;
  }
  \end{lstlisting}
  \caption{Dynamically detect aliasing case, and avoid by pushing another stack frame.}
  \label{lst:loopfixed}
\end{figure}

@@

\subsection{\texttt{malloc} will Alias by Default}
Acquiring dynamic memory at runtime is usually done by calling \texttt{malloc}, which takes a number of bytes to allocate as input, and returns a pointer to that area.
A typical malloc implementation uses different strategies for how to allocate memory depending on the \emph{size} of the request.
\begin{longitem}
  \item The ``regular'' heap area in virtual memory is used for smaller allocations. 
  Programs are initially given a relatively small heap, which the allocator can increase by calling \texttt{brk} or \texttt{sbrk} system calls.

  \item For larger allocations, it can be more efficient to use anonymous memory mapping by calling the Posix function \texttt{mmap} instead of growing the regular heap.
  One drawback of this is that it requires the memory to be zero initialized, and whether memory mapping is used or not is based on some cost heuristic employed by the allocator.
\end{longitem}
The heap section starts at a relatively low address (close to the data section), growing upwards.
Memory mapped chunks are placed towards the upper end of the virtual address space, closer to the stack.
Whether a memory allocation request is served by the heap or by memory mapping is therefore easy to determine just by looking at the actual pointer values returned.
While addresses in the regular heap can look something like 0x16e30a0 or 0x1723020, pointers returned by \texttt{mmap} are numerically much larger -- for example 0x7f0318a8f010 or 0x7f03105d2010.
This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
However, \texttt{mmap} has an interesting property in that allocations will always be page aligned.
The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of metadata at the beginning, therefore every memory mapped address end with 0x010.}
This behavior is often the worst case for functions that operate on two or more independent buffers, as we will see in the following sections.

\subsection{Aligned Sequential Access}
Many functions operate in a ``sliding window'' fashion; reading and writing to different buffers in some loop construction.
This type of program is potentially vulnerable to 4K aliasing, where the worst case will be when the read and write pointer addresses are aliased, continuously generating false conflicts.
As an example of this, consider a naive implementation of linear correlation shown in Figure \ref{lst:conv} (for simplicity, endpoints are not handled).
The function containins lots of interleaved loads and stores, iterating over two independent memory buffers in a tight loop.
We will see that the performance of this program greatly depends on the address alignment of each buffer, favoring memory addresses that are not closely aligned on the last 12-bits.

\begin{figure}[t]
  \centering
  \lstinputlisting[
    %float=t, 
    language=C, 
    %caption={Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.},
    %label={lst:conv},
    %frame=lines
  ]{bin/convolution-kernel.c}
  \caption{Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.}
  \label{lst:conv}
\end{figure}

By using an input size of $n=2^{20}$ (4 MiB in memory for each array), glibc's heap allocator ends up always choosing \texttt{mmap} to serve requests.
By default, even with address randomization enabled, both pointers will have the same address suffix of 0x010.
To analyse performance for different addresses, we manually insert padding to offset one of the memory mapped pointers.
This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset one of the arguments before calling the function.
Controlling the offset parameter, we can create environments where the inputs are some number of \texttt{sizeof(float)} bytes apart.
Allocating and managing these buffers at startup takes a non-neglegible amount of work.
The overhead can be masked by repeatedly invoking the \texttt{convolution} kernel after allocating and initializing all the inputs.
\begin{lstlisting}
    for (i = 0; i < k; ++i)
        convolution(n, input, output + offset);
\end{lstlisting}
An estimate of the actual cost of one invocation can be calculated by averaging over a number of repeated function calls, after subtracting the constant overhead of invoking it only once.

@@

\pgfplotstableread{bin/conv-default-o2.estimate.dat}{\convtabletwo}
\pgfplotstableread{bin/conv-default-o3.estimate.dat}{\convtablethree}
\begin{figure*}[t]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        title=\texttt{cc -O2},
        font=\footnotesize,
        xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
        cycle list name=black white,
        width=\textwidth/1.9, % Make it fit to text width side by side
        skip coords between index={20}{32} % Limit level of detail to fit page width nicely
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtabletwo ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwo ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=\texttt{cc -O3},
        font=\footnotesize,
        xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
        cycle list name=black white,
        width=\textwidth/1.9,
        skip coords between index={20}{32}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtablethree ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtablethree ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \label{fig:conv-default}
  \caption{Estimated cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Figure~\ref{lst:conv}. Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations. Input size $n=2^{20}$.}
\end{figure*}

% Table with detailed performance data, use estimated for O2 because it does not need as many columns (!)
\begin{table}[t]
  %\centering
  \label{tab:convstats}
  \renewcommand{\tabcolsep}{3pt} % Less padding between cells
  \tbl{Most interesting performance counters and correlation with cycle count for optimization O2, estimated cost accounting for constant overhead.}{
    \pgfplotstabletypeset[
      font=\footnotesize,
      int detect, % Output whole numbers for counter values
      col sep=comma,
      columns={Performance counter, Correlation, 0, 2, 4, 8},
      column type=r,
      columns/Performance counter/.style={
        string type, 
        column type=l,
        postproc cell content/.append code={
          \pgfkeysalso{@cell content=\perfctr{##1}}
        }
      },
      columns/Correlation/.style={
        fixed,
        fixed zerofill,
        precision=2
      },
      every head row/.style={
        output empty row,
        before row={\toprule
          Performance counter & $r$ & 0 & 2 & 4 & 8 \\
          },
        after row=\midrule,
      },
      every last row/.style={after row=\bottomrule}
    ]{bin/conv-default-o2.estimate.csv}
  }
\end{table}

@@

\begin{table*}[t]
  \centering
  \label{tab:mallocompare}
  \renewcommand{\tabcolsep}{4pt} % Less padding between cells
  \tbl{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.}{
    \pgfplotstabletypeset[
      font=\footnotesize,
      col sep=comma,
      string type,
      columns={Allocation, 64, 5120, 65536, 1048576, 4194304},
      column type=r,
      columns/Allocation/.style={
        string type, 
        column type=l
      },
      every head row/.style={
        output empty row,
        before row={\toprule
           & 64 B & 5,120 B & 65,536 B & 1,048,576 B & 4,194,304 B \\
          },
        after row=\midrule,
      },
      every odd row/.style={after row=\midrule},
      every last row/.style={after row=\bottomrule}
    ]{bin/malloc-comparison.csv}
  }
\end{table*}

@@

\paragraph{Manually adjust address offsets}
In some cases it might make sense to explicitly control the memory addresses used, for example forcing some fixed relative offset between input and output pointers.
This can be achieved by exploiting the fact the \texttt{mmap} is in fact guaranteed to be placed at a page boundary, and use that directly instead of \texttt{malloc}.
The following approach can be used to make an anonymous memory mapping with offset \texttt{d} bytes away from page alignment.
\begin{lstlisting}[breaklines=true]
    mmap(NULL, (n + d), PROT_READ | PROT_WRITE, MAP_PRIVATE | MAP_ANONYMOUS, -1, 0) + d;
\end{lstlisting}
The same pointer difference must of course by subtracted before unmapping the memory.
